%%% Time-stamp: <2022-05-13 20:06:05 vladimir>
%%% Copyright (C) 2019-2022 Vladimir G. Ivanović
%%% Author: Vladimir G. Ivanović <vladimir@acm.org>
%%% ORCID: https://orcid.org/0000-0002-7802-7970
\chapter{Acknowledgments}

\begin{comment}
My debts are many.
  To Diane Ravitch, whose fierce and intellectually rigorous focus on the inequities of public education, I owe my much weaker commitment to changing public policy so that every child gets the education they need to flourish. To Robin Avelar LaSalle, with whom I lost every (friendly) education argument we had during our first dinner, I owe my acceptance into the Ed.D. program at San José State University. Her letter of recommendation was, I'm sure, critical to my getting into the Ed.D. program. To Roxana Marachi, my dissertation advisor, put simply, I owe my doctorate. I cannot imagine having a better committee chair and I hope to accomplish half of what she has. To Brad Porfilio, I owe much gratitude for his support and leadership of the Ed.D. program. To my long suffering wife, Carol, and to our two children, I owe so much for their tolerance of my quirks, my absence, and the vast number of books that creep into every corner of our house. To the professors in the Ed.D. program who did their very best to instill in me the requisites of becoming an education leader, I owe many thanks. Lastly, to Cohort 6 (Class of '22), and especially to Gigi Carunungan and Candice Nance, I thank you for your unwavering support.

It goes without saying that I am solely responsible for any errors or omissions in this dissertation.
\end{comment}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "Rocketship_Education:_An_Exploratory_Public_Policy_Case_Study"
%%% End:
