%%% Time-stamp: <2024-01-12 10:02:37 vladimir>
%%% Copyright (C) 2019-2024 Vladimir G. Ivanović
%%% Author: Vladimir G. Ivanović <vladimir@acm.org>
%%% ORCID: https://orcid.org/0000-0002-7802-7970
\chapter{Acknowledgments}\indent%

My debts are many.
  
\begin{comment}

To Diane Ravitch, whose fierce and intellectually rigorous focus on the inequities of public education, I owe my much weaker commitment to changing public policy so that every child gets the education they need to flourish. To Robin Avelar LaSalle, with whom I lost every (friendly) education argument we had during our first dinner, I owe my acceptance into the Ed.D. program at San José State University. Her letter of recommendation was, I'm sure, critical to my getting into the Ed.D. program. Her lifelong focus on righting the inequities in educational policy in California has inspired me. To Roxana Marachi, my dissertation advisor, to whom, put simply, I owe my doctorate. I truly cannot imagine a better committee chair and I hope to accomplish half of what she has. To Gordon Lafer, a committee member, and to XXXX, also a committee member, thank you for pointing out infelicities in my writing and thinking. To Arnie Danzig, Brad Porfilio, and Fergie Revera, Ed.D. Program Directors, I owe much gratitude for their support at critical junctures in my journey to a doctorate. To my long suffering wife, Carol, and to our two children, I owe so much for their tolerance of my quirks, not to mention the vast number of books that crept into every corner of our house. To the professors in the Ed.D. program who did their very best to instill in me the requisites of becoming an education leader, I owe much. To Jennifer Carlstrom, who was the first to guide me on my journey through the thicket of educational policy in California, thank you for getting me started. To Jeff Baier, Sandra McGonagle, and Carrie Bosco (herself an Ed.D. graduate), all senior administrators in the Los Altos School District, thank you for listening graciously to my many suggestions for improvement from someone who has never actually run a school district. Lastly, to Cohort 6 (Class of '22), and especially to Gigi Carunungan and Candice Nance, thank you many times over for your unwavering support (and ceaseless pestering, "Is it done yet?").

Nothing I claim as my own was written by an artificial intelligence program. I am solely responsible for any and all errors in this dissertation.

\end{comment}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "Rocketship_Education-An_Exploratory_Public_Policy_Case_Study"
%%% End:
