%%% Time-stamp: <2023-06-13 16:25:06 vladimir>
%%% Copyright (C) 2019-2023  Vladimir G. Ivanović
%%% Author: Vladimir G. Ivanović <vladimir@acm.org>
%%% ORCID: https://orcid.org/0000-0002-7802-7970
\chapter{Colophon}
\small\OnehalfSpacing%

Two versions of this dissertation are available. The contents of both are identical; only the formatting differs.
\begin{enumerate}[nosep]
  \item \url{https://leonora.org}\\\noindent%
  This version, the readable version, has been formatted according to standard typographic practice which emphasizes  readability, and it is hoped, good taste and elegance. 
  \item \url{https://sjsu.edu}\\\noindent%
  This version, the official, submitted version, has been formatted to meet the SJSU Ed.D. doctoral dissertation guidelines. You should read this version only under duress.
\end{enumerate}

This dissertation was created nearly entirely with free, open source programs.\footnote{Two closed source, proprietary programs were used to manipulate PDF files, themselves proprietary programs until recently. These programs could have been replaced with FOSS programs, but the results would not have been as high quality.} The typefaces, the text editor, the markup language, the reference manager, the operating system, and many utilities are all FOSS (free, open source software).
\begin{center}
  \textbf{\adforn{21}\qquad\adforn{21}\qquad\adforn{11}\qquad\adforn{49}\qquad\adforn{49}}
\end{center}
The body and headings were set in 12pt Alegreya. The Alegreya family of serif \& sans serif typefaces was designed by Juan Pablo del Peral of Huerta Tipográfica in 2011 and immediately won praise and awards. It is a classic Renaissance typeface, a kind that was first developed in the fourteenth and fifteenth centuries in northern Italy. It comes in Regular, Medium, Bold and Black weights, all of which are available in Roman and Italic styles. There is a full set of Greek and Cyrillic letters as well as Latin small caps. All have a full set of ligatures, and Old Style, and Lining numerals. Notably, all the numerals share the same width so they line up regardless of which style is being used. (Multiplication using Roman numerals, anyone?) If any criticism can be leveled against the Alegreya superfamily, it is that the family doesn't include display sizes and doesn't contain swash characters. Otherwise it is nearly perfect.
\begin{center}
  \textbf{\adforn{21}\quad\quad\adforn{21}\quad\quad\adforn{11}\quad\quad\adforn{49}\quad\quad\adforn{49}}
\end{center}
The programs \TeX{} \& \LaTeX{}  and the document class \texttt{memoir} were used to format this dissertation. \LaTeX  was created by Leslie Lamport as a user-friendly version of one of the first digital typesetting systems, \TeX . \TeX  is one of the masterpieces of computer programming whose author, Donald Knuth, won the Turing Award in 1974. It is a testament to Knuth's brilliance as both a mathematician and a programmer that \TeX  is still in use more than four decades later and arguably has no peers when it comes to typesetting complex mathematics and scientific material. It is, however, awkward to use and hard to learn. Fortunately, Leslie Lamport wrapped \TeX  in a macro system, \LaTeX , which was orders of magnitude easier to use than \TeX  itself.
LaTeX is extraordinarily flexible because there are thousands of packages which implement specialized tasks. Currently, CTAN (the Comprehensive TeX Archive Network) has just shy of 6000 packages which can be downloaded. One of those packages implements the class \texttt{memoir} that was used here. It was written by Peter Wilson, and released in 2001. (I'm listed as a contributor to \texttt{memoir}, but in truth I really only just corrected some minor typos.)
\newpage
\begin{center}
  \textbf{\adforn{21}\quad\quad\adforn{21}\quad\quad\adforn{11}\quad\quad\adforn{49}\quad\quad\adforn{49}}
\end{center}
Wilson's muse is Robert Bringhurst, author of \textit{The Elements of Typographic Style}, which some consider to be the definitive book on typography and book design. It is certainy the most elegant. The package \texttt{memoir} would undoubtedly meet with Bringhurst's approval.  The class \texttt{memoir} provides in one package nearly everything a person needs to produce ``beautiful books'' (Knuth's words). Although creating a bibliography, glossary, and an index are possible in \texttt{memoir}, specialized packages are normally used instead of the built-in ones supplied by \texttt{memoir}.
\begin{center}
  \textbf{\adforn{21}\quad\quad\adforn{21}\quad\quad\adforn{11}\quad\quad\adforn{49}\quad\quad\adforn{49}}
\end{center}
Zotero is a program to manage and maintain a bibliographic database and to provide citations on demand. It, along with the text editor GNU Emacs (``an operating system disguised as an editor'') and the package \texttt{reftex}, cooperate with \texttt{memoir} to provide a complete system for writing scholarly papers, theses, reports, and dissertation.
\begin{center}
  \textbf{\adforn{21}\quad\quad\adforn{21}\quad\quad\adforn{11}\quad\quad\adforn{49}\quad\quad\adforn{49}}
\end{center}
All of these program run on Linux, a version of Unix using a particular distribution, Arch Linux, one of many. It is notable that GNU Linux, GNU Emacs, \TeX  and \LaTeX are all programs that are decades old, have never been replaced or superseded, are constantly being improved, and are actively used. They share a common set of characteristics: their fundamental architecture is sound, extensibility is a core feature, and they and thousands of specialized packages are freely available. I predict that iPhones will be the faintest of memories when GNU Unix, GNU Emacs, and \LaTeX start to fade from view.
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "Rocketship_Education-An_Exploratory_Public_Policy_Case_Study"
%%% End:
