% Time-stamp: <2023-09-02 12:49:07 vladimir>
% Copyright (C) 2019-2023 Vladimir G. Ivanović
% Author: Vladimir G. Ivanović <vladimir@acm.org>
% ORCID: https://orcid.org/0000-0002-7802-7970
% arara: lualatex

\input{preamble}

\addbibresource{References.bib}

\pretitle{}
\title{%
  \begin{center}
    \normalfont%
    ROCKETSHIP EDUCATION:\\
    AN EXPLORATORY PUBLIC POLICY CASE STUDY
  \end{center}
} % SJSU EDD Guidelines Checklist #17
\posttitle{}

\date{}

\preauthor{%
  \begin{center}
    \normalfont%
    \bigskip\bigskip\bigskip\bigskip\bigskip
    A Dissertation Presented to\\
    The Faculty of the Connie L. Lurie College of Education\\
    San José State University\\
    \bigskip\bigskip\bigskip\bigskip\bigskip
    In Partial Fulfillment of the Requirements for the Degree\\
    Doctor of Education\\
  \end{center}
}

\author{%
  \begin{center}
    by\\
    Vladimir Gresham Ivanović \protect\orcidlink{0000–0002–7802–7970}\\\bigskip\bigskip
  \end{center}
}

\postauthor{%
  \begin{center}
    December 2023\\
    \vspace{1.5in}
    Version \version{} created on \DTMnow%
  \end{center}
}

\hypersetup{%
  pdfinfo={%
    Title={Rocketship Education: An Exploratory Public Policy Case Study},
    Author={Vladimir Gresham Ivanović},
    Language={en-US},
    Subject={An Ed.D. dissertation that examines the finances of Rocketship Education, a charter school management organization, using a public policy lens.},
    Keywords={Rocketship Education, charter management organization, CMO, charter finances, education public policy},
    Version={\version}
  }
}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\frontmatter{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

\thispagestyle{empty}           % for title page
\pagestyle{empty}               % for copyright, signature, & abstract pages

\input{Copyright}
\newpage\input{Signature}\clearpage
\mbox{}                         % Force an empty page
\cleardoublepage%
\input{Abstract}

\pagestyle{plain}               % start displaying folios, but using Roman numerals until the main matter.

% \newpage\input{Acknowledgments}
\OnehalfSpacing%
\newpage{\tableofcontents*}
\newpage\listoftables%
\newpage\listoffigures%
\newpage\input{Abbreviations}
\clearforchapter\input{Glossary}
\newpage\listoftodos[List of things still to do]%
\DoubleSpacing%

%\clearforchapter\input{Epigraph}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\mainmatter{}
% \linenumbers%
\todo{Review tax law changes that affect charter school operators to see which have the most effect.}
\todo{Make sure the organization of my dissertation is crystal clear to readers.}
\todo{Make sure the limitations of my study are spelled out.}
\todo{Fix the capitalization of references according to SJSU \& APA.}
\todo{Check \& recheck all tiny URLs.}
\todo{Make sure citations are correct.}
\todo{Make sure that all biber/biblatex annotations in References.tex are accounted for.} 
\todo{Format preamble (mylatexformat) to speed compilation.}
\todo{Add "Chapter" and number to chapter headings.}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearforchapter\input{Introduction}                  % Chapter 1
\clearforchapter\input{Literature_Review}             % Chapter 2
\clearforchapter\input{Methods}                       % Chapter 3
\clearforchapter\input{Findings}                      % Chapter 4
\clearforchapter\input{Discussion}                    % Chapter 5
\newpage%

\setlength{\afterchapskip}{0.8\onelineskip}

\OnehalfSpacing%
\setlength{\bibitemsep}{\onelineskip}
\printbibliography{}

\DoubleSpacing%
\appendix%
\part{Appendices}
\begin{appendices}
\chapterstyle{appdx_sjsu}
\clearforchapter\input{School_Financing}
\clearforchapter\input{Santa_Clara_County_Charter_Schools}
\clearforchapter\input{Rocketship_Properties}

\clearforchapter\input{Consolidated_Financial_Position_Years_2010-2022}
\clearforchapter\input{Consolidated_Activities_Years_2010-2022}
\clearforchapter\input{Consolidated_Cash_Flows_Years_2010-2022}
\clearforchapter\input{Consolidated_Functional_Expenses_Years_2019-2022}

\clearforchapter\input{Debt_2008-2022.tex}
\end{appendices}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\backmatter{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearforchapter{\small\printindex}
% \linenumbers%
% \clearforchapter\input{Colophon}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "Rocketship_Education-An_Exploratory_Public_Policy_Case_Study"
%%% End:
